---
title: "Data Visualization: looking at data"
author: "Kylie Ariel Bemis"
date: "9/17/2020"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r tidyverse, message=FALSE}
library(tidyverse)
```

# Looking at data

## Why do we look at data?

Here's a dinosaur dataset:

```{r dino}
library(datasauRus)

datasaurus_dozen %>%
  filter(dataset == "dino") %>%
  ggplot(aes(x=x, y=y)) +
  geom_point() +
  labs(x="x", y="y") +
  theme_minimal()
```

Here's 12 more datasets with identical summary statistics:

```{r dino-dozen-stats}
datasaurus_dozen %>% 
  group_by(dataset) %>% 
  summarize(
    mean_x    = mean(x),
    mean_y    = mean(y),
    std_dev_x = sd(x),
    std_dev_y = sd(y),
    corr_x_y  = cor(x, y))
```

But plotting them reveals quite a difference!

```{r dino-dozen-plots}
datasaurus_dozen %>%
  filter(dataset != "dino") %>%
  ggplot(aes(x=x, y=y, colour=dataset)) +
  geom_point() +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(~dataset, ncol=3)
```

## Example data

### Gapminder

A small, tidied subset of the Gapminder data on life expectancy, GDP per capita, and population by country.

```{r gapminder}
library(gapminder)

gapminder
```

Source: http://www.gapminder.org/data/

### Fuel economy

A small, tidied subset of fuel economy data from 1999 to 2008 for 38 popular models of cars, provided by the EPA.

```{r mpg}
library(ggplot2)

mpg
```

Source: http://fueleconomy.gov/

### Simulated

Symmetric normal, left-skew, right-skew, and uniform data.

```{r sim-data}
set.seed(1)
sim <- data.frame(x1=rnorm(1000),
                  x2=rlnorm(1000, sdlog=0.5),
                  x3=-rlnorm(1000, sdlog=0.5),
                  x4=runif(1000))
```

## Single-variable graphics

### Histogram

```{r histogram-gapminder-1}
ggplot(gapminder, aes(x=lifeExp)) +
  geom_histogram() +
  labs(title="Distribution of life expectancy (bins = 30)",
       x="Life expectancy (years)") +
  theme_minimal()
```

```{r histogram-gapminder-2}
ggplot(gapminder, aes(x=lifeExp)) +
  geom_histogram(bins = 50) +
  labs(title="Distribution of life expectancy (bins = 50)",
       x="Life expectancy (years)") +
  theme_minimal()
```

```{r histogram-gapminder-3}
ggplot(gapminder, aes(x=lifeExp)) +
  geom_histogram(binwidth = 1) +
  labs(title="Distribution of life expectancy (binwidth = 1 year)",
       x="Life expectancy (years)") +
  theme_minimal()
```

```{r histogram-sim-1}
ggplot(sim, aes(x=x1)) +
  geom_histogram() +
  labs(title="Symmetric") +
  theme_minimal()
```

```{r histogram-sim-2}
ggplot(sim, aes(x=x2)) +
  geom_histogram() +
  labs(title="Right skew") +
  theme_minimal()
```

```{r histogram-sim-3}
ggplot(sim, aes(x=x3)) +
  geom_histogram() +
  labs(title="Left skew") +
  theme_minimal()
```

### Density plot

```{r density-gapminder-1}
ggplot(gapminder, aes(x=lifeExp)) +
  geom_density(fill="darkcyan") +
  labs(title="Life expectancy (binwidth = 1 year)",
       x="Life expectancy (years)") +
  theme_minimal()
```

```{r density-sim-1}
ggplot(sim, aes(x=x1)) +
  geom_density(fill="darkcyan") +
  labs(title="Symmetric") +
  theme_minimal()
```

```{r density-sim-2}
ggplot(sim, aes(x=x2)) +
  geom_density(fill="darkcyan") +
  labs(title="Right skew") +
  theme_minimal()
```

```{r density-sim-3}
ggplot(sim, aes(x=x3)) +
  geom_density(fill="darkcyan") +
  labs(title="Left skew") +
  theme_minimal()
```

### Box plot

```{r boxplot-gapminder-1}
ggplot(gapminder, aes(x=lifeExp)) +
  geom_boxplot() +
  lims(y=c(-1, 1)) +
  labs(title="Distribution of life expectancy",
       x="Life expectancy (years)") +
  theme_minimal()
```

```{r boxplot-sim-1}
ggplot(sim, aes(x=x1)) +
  geom_boxplot() +
  lims(y=c(-1, 1)) +
  labs(title="Symmetric") +
  theme_minimal()
```

```{r boxplot-sim-2}
ggplot(sim, aes(x=x2)) +
  geom_boxplot() +
  lims(y=c(-1, 1)) +
  labs(title="Right skew") +
  theme_minimal()
```

```{r boxplot-sim-3}
ggplot(sim, aes(x=x3)) +
  geom_boxplot() +
  lims(y=c(-1, 1)) +
  labs(title="Left skew") +
  theme_minimal()
```

### Bar plot

```{r barplot-mpg-1}
ggplot(mpg, aes(x=class)) +
  geom_bar() +
  labs(title="Class of car") +
  theme_minimal()
```

```{r barplot-mpg-2}
ggplot(mpg, aes(x=manufacturer)) +
  geom_bar() +
  labs(title="Manufacturer representation") +
  coord_flip() +
  theme_minimal()
```

### Pie chart

```{r pie-mpg-1}
ggplot(mpg, aes(x=1, fill=class)) +
  geom_bar() +
  labs(title="Class of car") +
  coord_polar("y") +
  theme_minimal()
```

### Q-Q plot

```{r qq-sim-1}
ggplot(sim, aes(sample=x1)) +
  geom_qq() +
  labs(title="Symmetric") +
  theme_minimal()
```

```{r qq-sim-2}
ggplot(sim, aes(sample=x2)) +
  geom_qq() +
  labs(title="Right skew") +
  theme_minimal()
```

```{r qq-sim-3}
ggplot(sim, aes(sample=x3)) +
  geom_qq() +
  labs(title="Left skew") +
  theme_minimal()
```

## Multi-variable graphics

### Scatter plot

```{r scatter-mpg-1}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  labs(title="Larger engines are less efficient",
       x="Engine size (liters)", y="Highway fuel economy (mpg)") +
  theme_minimal()
```

```{r scatter-mpg-2}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  geom_smooth() +
  labs(title="Larger engines are less efficient",
       x="Engine size (liters)", y="Highway fuel economy (mpg)") +
  theme_minimal()
```

```{r scatter-mpg-3}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_jitter() +
  geom_smooth() +
  labs(title="Larger engines are less efficient",
       x="Engine size (liters)", y="Highway fuel economy (mpg)") +
  theme_minimal()
```

### 2D density plot

```{r density2d-mpg-1}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_density2d_filled() +
  labs(title="Larger engines are less efficient",
       x="Engine size (liters)", y="Highway fuel economy (mpg)") +
  theme_minimal()
```

### Box plots

```{r boxplots-mpg-1}
ggplot(mpg, aes(x=class, y=hwy)) +
  geom_boxplot() +
  labs(title="Fuel economoy for different car classes",
       x="Type of car", y="Highway fuel economy (mpg)") +
  theme_minimal()
```

```{r boxplots-mpg-2}
ggplot(mpg, aes(x=manufacturer, y=hwy)) +
  geom_boxplot() +
  labs(title="Fuel economoy for different manufacturers",
       x="Manufacturer", y="Highway fuel economy (mpg)") +
  coord_flip() +
  theme_minimal()
```

### Histograms

```{r histograms-mpg-1}
ggplot(mpg, aes(x=hwy, color=class)) +
  geom_freqpoly(bins=20) +
  labs(title="Fuel economoy for different car classes",
       x="Highway fuel economy (mpg)") +
  theme_minimal()
```

### Density plots

```{r density-mpg-1}
ggplot(mpg, aes(x=hwy, color=class, fill=class)) +
  geom_density(alpha=1/3) +
  labs(title="Fuel economoy for different car classes",
       x="Highway fuel economy (mpg)") +
  theme_minimal()
```

### Bar plots

```{r bars-mpg-1}
ggplot(mpg, aes(x=class, fill=as.factor(year))) +
  geom_bar(position="dodge") +
  scale_fill_discrete(name="Year") +
  labs(title="Breakdown of car types by year") +
  theme_minimal()
```

```{r bars-mpg-2}
ggplot(mpg, aes(x=class, fill=as.factor(year))) +
  geom_bar(position="stack") +
  scale_fill_discrete(name="Year") +
  labs(title="Breakdown of car types by year") +
  theme_minimal()
```

## Faceting

### Faceting by one variable

```{r facets-mpg-1}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~year) +
  labs(title="Fuel efficiency vs Engine size by Year",
       x="Engine size (liters)", y="Highway fuel economy (mpg)") +
  theme_minimal()
```

```{r facets-gapminder-1}
ggplot(gapminder, aes(x=gdpPercap, y=lifeExp)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~year) +
  scale_x_log10() +
  labs(title="Life expectancy increases with GDP per capita",
       x="GDP per capita (log10 dollars)", y="Life expectancy (years)") +
  theme_minimal()
```

### Faceting by two variables

```{r facets-gapminder-2}
gapminder %>%
  filter(continent != "Oceania") %>%
  ggplot(aes(x=gdpPercap, y=lifeExp)) +
  geom_point() +
  geom_smooth() +
  facet_grid(continent~year) +
  scale_x_log10() +
  labs(title="Life expectancy increases with GDP per capita",
       x="GDP per capita (log10 dollars)", y="Life expectancy (years)") +
  theme_minimal()
```
