---
title: "Introduction to ggplot2: grammar of graphics"
author: "Kylie Ariel Bemis"
date: "9/18/2020"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A Layered "Grammer of Graphics"

Components of a plot:

- Default dataset and set of mappings from variables to aesthetics

- One or more layers, each having:

    - A geometric object
    
    - A statistical transformation
    
    - A position adjustment
    
    - (Optional) A dataset
    
    - (Optional) A set of aesthetic mappings
    
- A scale for each mapped aesthetic

- A coordinate system

- A facet specification


## A simple example (revisited)

```{r}
simple <- data.frame(A=c(2, 1, 4, 9),
                    B=c(3, 2, 5, 10),
                    C=c(4, 1, 15, 80),
                    D=c('a', 'a', 'b', 'b'))
simple
```

We wish to create a scatter plot of A versus C, using shape for D.

What are:

- The default dataset and aesthetic mappings?

- The layers?

    - The geometric object?
    
    - The statistical transformation?
    
    - The position adjustment?
    
- The scales for the mapped aesthetics?
    
- The coordinate system?

- The facet specification?

### Plotting with `ggplot2` syntax

```{r simple}
library(ggplot2)

ggplot(data=simple,
       mapping=aes(x=A, y=C, shape=D)) +
  layer(geom="point",
        stat="identity",
        position="identity") +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_shape_discrete() +
  coord_cartesian() +
  facet_null()
```

### Faceting with `ggplot2` syntax

```{r simple-facet}
ggplot(data=simple,
       mapping=aes(x=A, y=C, shape=D)) +
  layer(geom="point",
        stat="identity",
        position="identity") +
  scale_x_continuous() +
  scale_y_continuous() +
  coord_cartesian() +
  facet_wrap(~D)
```

## Sensible defaults

A number of these can be considered sensible defaults:

- For `geom="point"`, use `stat="identity"` unless otherwise specified

- For `geom="point"`, use `position="identity"`  unless otherwise specified

- Cartesian coordinate system by default

- No facets unless explicitly specified

## The usual way -- the concise `ggplot2` syntax

```{r simpler}
ggplot(simple, aes(x=A, y=C, shape=D)) + 
  geom_point()
```

## Every `geom` has a default `stat`

- Scatter plot - `geom_point`
    - Identity - `stat_identity`

- Line plot - `geom_line`
    - Identity - `stat_identity`

- Box plot - `geom_boxplot`
    - Boxplot (five summary statistics + outliers) - `stat_boxplot`

- Histogram - `geom_histogram`
    - Binning - `stat_bin`

- Bar plot - `geom_bar`
    - Count - `stat_count`

These can always be changed!

# Example: Fuel Economy in Cars

```{r}
mpg
```

## Mapping aesthetics

Plot engine size versus highway miles per gallon

```{r mpg-1}
ggplot(data = mpg,
       mapping = aes(x=displ, y=hwy)) + 
  geom_point()
```

Map class to color

```{r mpg-2}
ggplot(data = mpg,
       mapping = aes(x=displ,
                     y=hwy,
                     color=class)) + 
  geom_point()
```

Map number of cylinders to size

```{r mpg-3}
ggplot(data = mpg,
       mapping = aes(x=displ,
                     y=hwy,
                     color=class,
                     size=cyl)) + 
  geom_point()
```

## Faceting

Facet by drive type (front/rear/4-wheel)

```{r mpg-4}
ggplot(data = mpg,
       mapping = aes(x=displ,
                     y=hwy,
                     color=class,
                     size=cyl)) + 
  geom_point() + facet_grid(~drv)
```

## Annotations

Add plot annotations such as axis labels, title, theme, etc.

```{r mpg-5}
ggplot(data = mpg,
       mapping = aes(x=displ,
                     y=hwy,
                     color=class,
                     size=cyl)) + 
  geom_point() + facet_wrap(~drv) +
  labs(x="Engine Size (Liters)",
       y="Highway Miles per Gallon",
       title="Larger engines are less fuel-efficient") +
  theme_minimal()
```

Good plots have informatively labeled axes (with units!) and an explanatory title.

This rule may be broken during exploratory data analysis, but make sure to annotate your plots for presentation!

## Additional `geom`s

What if we want to add a smooth line?

```{r mpg-smooth-1}
ggplot(data = mpg,
       mapping = aes(x=displ, y=hwy)) + 
  layer(geom="point",
        stat="identity",
        position="identity") +
  layer(geom="line",
        stat="smooth",
        position="identity",
        params=list(method="auto", formula=y~x, se=TRUE))
```

Or more simply, use `geom_smooth`

```{r mpg-smooth-2}
ggplot(data = mpg,
       mapping = aes(x=displ, y=hwy)) + 
  geom_point() + geom_smooth()
```


### `geom`s in `ggplot2`

- Geoms in `ggplot2` represent a layer with a set of defaults

    - Geometric object

    - Statistical transformation
    
    - Position adjustment
    
- Geoms in `ggplot2` are shortcuts for potentially complex layers

- Geoms in `ggplot2` are sometimes redundant with other geoms

### Histograms and bar plots

```{r mpg-hist-1}
ggplot(data = mpg, mapping = aes(x=class)) +
  geom_bar()
```

```{r mpg-hist-2}
ggplot(data = mpg, mapping = aes(x=hwy)) +
  geom_histogram()
```

Why doesn't histogram use the "bar" geom?

----

We can rewrite both explicitly to use `geom="bar"` and different stats.

```{r mpg-bar-3}
ggplot(data = mpg, mapping = aes(x=class)) +
  layer(geom="bar",
        stat="count",
        position="identity")
```

```{r mpg-hist-4}
ggplot(data = mpg, mapping = aes(x=hwy)) +
  layer(geom="bar",
        stat="bin",
        position="identity")
```

----

We could also use `geom_bar` for both

We can make a histogram with `geom_bar` by overwriting the default stat:

```{r mpg-hist-5}
ggplot(data = mpg, mapping = aes(x=hwy)) +
  geom_bar(stat="bin")
```

### Boxplots

- Boxplots are unique `geom` with a unique `stat`

- They plot the five-number summary + outliers

- Here we plot side-by-side boxplots for highway mpg by class

```{r mpg-boxplot}
ggplot(data = mpg, mapping = aes(x=class, y=hwy)) +
  geom_boxplot()
```

# A template for `ggplot2` plotting

We can develop a simple template for creating plots in `ggplot2`:

```{}
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + 
  <GEOM_FUNCTION>()
```

You will notice this is slightly different from the template that appears in the _R for Data Science_ -- how and why?

# Layers in `ggplot2`

As we've already seen, it is possible to overlay different geoms:

```{r defaults-1}
ggplot(data = mpg,
       mapping = aes(x=displ, y=hwy, color=drv)) + 
  geom_point() + geom_smooth()
```

In this case, both geoms inherit the default data and aesthetic mappings.


## Different aesthetics

However, suppose we want to use different aesthetics for each geom.

We can either give each layer its own aesthetic mapping:

```{r defaults-3}
ggplot(data = mpg) + 
  geom_point(mapping=aes(x=displ, y=hwy, color=drv)) + 
  geom_smooth(mapping=aes(x=displ, y=hwy))
```

Or we can supply a default aesthetic and override it when necessary:

```{r defaults-4}
ggplot(data = mpg,
       mapping=aes(x=displ, y=hwy)) + 
  geom_point(mapping=aes(color=drv)) + 
  geom_smooth()
```

## Different data

We can also specify different datasets for each layer, or allow them to inherit from the default dataset.

```{r defaults-5}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = drv)) + 
  geom_smooth(data = mpg[mpg$drv == "f",])
```

## Overplotting: rounded values

Consider the following plot:

```{r overplot-1}
ggplot(data = mpg,
       mapping = aes(x=displ, y=hwy)) + 
  geom_point()
```


Notice how all of the data points are on neat lines?

### Solution: jitter

```{r jitter}
ggplot(data = mpg,
       mapping = aes(x=displ, y=hwy)) + 
  geom_point(position="jitter")
```

## Overplotting: too much data

```{r overplot-2}
ggplot(diamonds, mapping=aes(x=carat, y=price)) +
  geom_point()
```

### Solution: transparency

```{r alpha}
ggplot(diamonds, mapping=aes(x=carat, y=price)) +
  geom_point(alpha=1/100)
```

# Geoms and stats in `ggplot2`

## Geom + stat recipes

Geom + stat define many common types of statistical plots

What are these geom + stat combinations?

- geom = "point", stat = "identity"

    - Scatterplot

- geom = "bar", stat = "count"

    - Bar plot

- geom = "bar", stat = "bin"

    - Histogram

- geom = "point", stat = "qq"

    - Quantile-quantile plot

- geom = "boxplot", stat = "boxplot"

    - Boxplot

## Default stats

What are default stats for these geoms?

- `geom_point`

    - `stat_identity`

- `geom_line`

    - `stat_identity`

- `geom_bar`

    - `stat_count`

- `geom_boxplot`

    - `stat_boxplot`

## Remember: geoms are shortcuts

Each geom function represents a set of defaults for:

- A geometric object

- A statistical transformation

- A position adjustment

That means some ggplot2 'geoms' represent the same geometric object w/ different statistical transformations.

What are some examples of this?

- geom = "bar" : geom_bar, geom_histogram

- geom = "point" : geom_point, geom_qq

Are there any ggplot2 'geoms' that represent the same geometric object and statistical transformation but with different position adjustments?

- geom = "point" : geom_point, geom_jitter

### How can we inspect what the defaults are?

1. Check the help page (`?geom_point`, `?geom_qq`)

2. Check the body and formal arguments of the function

```{r}
geom_histogram
```

# More plotting with `ggplot2`

Minimally, when plotting with `ggplot2`:

- Initialize a with plot `ggplot()` and (optionally):

    - A default dataset
    
    - A default set of aesthetic mappings

- Add at least one layer with a geometric object and default statistical transformation with `geom_xxx()` and (optionally):

    - A dataset
    
    - A set of aesthetic mappings

## Building a plot incrementally

With `ggplot2`, plots can be assigned to a variable to be modified and plotted later.

A sequence of `ggplot2` functions won't be plotted (yet) if assigned to a variable.

```{r}
g <- ggplot(data=mpg)
```

---

```{r}
g + geom_point(mapping=aes(x=displ, y=hwy))
```

---

```{r}
g + geom_jitter(mapping=aes(x=displ, y=hwy))
```



## Using `stat` functions

Sometimes it makes more sense to think about the statistical transformation you want to display visually first.

Because each 'stat' also have a default 'geom' in `ggplot2`, you can use the `stat_xxx` functions for plotting as well.

What do you think the default geoms are for these stats?

- `stat_identity`

    - `geom_point`

- `stat_count`

    - `geom_bar`

- `stat_bin`

    - `geom_bar`

- `stat_qq`

    - `geom_point`
    
- `stat_boxplot`

    - `geom_boxplot`

---

```{r}
ggplot(data=mpg) + stat_identity(mapping=aes(x=displ, y=hwy))
```

---

```{r}
ggplot(data=mpg) + stat_count(mapping=aes(x=drv))
```

---

```{r}
ggplot(data=mpg) + stat_bin(mapping=aes(x=hwy))
```

---

```{r}
ggplot(data=mpg) + stat_qq(mapping=aes(sample=hwy))
```

---

```{r}
ggplot(data=mpg) + stat_boxplot(mapping=aes(x=drv, y=hwy))
```

## When to use `stat_xxx` instead of `geom_xxx` functions?

- Using the `stat_xxx` functions to specify plots is not as common in `ggplot2` as using the `geom_xxx` functions

- However, in some cases, the statistical transformation you want to visualize is more readily obvious than the geometric object

- Adjusting parameters for complex statistical transformations (e.g., smoothed trend lines) may be easier using the `stat_xxx` function

- Use whichever is more intuitive for you

**Most plot types can be defined by a geom + stat.**

# Computed variables

Many statistical transformations calculate values to be plotted.

A basic example are histograms and bar plots.

By default, both `geom_histogram` and `geom_bar` plot counts.

What if we want to plot the density of each bin?

What if we want to plot the proportion for each group?

---

Use `..density..` to plot densities in a histogram.

```{r hist-1}
ggplot(data=mpg) + 
  geom_histogram(mapping=aes(x=hwy))
```

```{r hist-2}
ggplot(data=mpg) + 
  geom_histogram(mapping=aes(x=hwy, y=..density..))
```

---

Use `..prop..` to plot proportions in a bar plot.

```{r bar-1}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv))
```

```{r bar-2}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv, y=..prop..))
```

What went wrong above?

## Finding computed variables

How do we find out exactly what is calculated?

Check the documentation:

```{r eval=FALSE}
?geom_bar
?stat_count
```

----

**"Bar charts"**

Description

> There are two types of bar charts: geom_bar makes the height of the bar proportional to the number of cases in each group...

Computed variables

- count

    - number of points in bin
    
- prop

    - groupwise proportion

---

Set the `group` aesthetic to a constant to calculate correct proportions.

```{r bar-3}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv, y=..prop.., group=1))
```

---

Does it matter what `group` is as long as it's a constant?

```{r}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv, y=..prop.., group=2))
```

Will this work?

```{r}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv, y=..prop.., group=drv))
```

```{r}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv, y=..prop.., group="foo"))
```

---

Can we replicate the original behavior while specifying `y` explicitly?

```{r}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv)) +
  labs(title="Implicit y")
```

versus

```{r}
ggplot(data=mpg) + 
  geom_bar(mapping=aes(x=drv, y=..count..)) +
  labs(title="Explicit y")
```

## More on calculated variables

- What does `stat_smooth` calculate?

    - What geom does it use?
    
    - Why doesn't it use `geom_line`?

```{r smooth, eval=FALSE}
ggplot(data=mpg) + stat_smooth(aes(x=displ,
                                   y=hwy),
                               geom="smooth")
```

versus

```{r line, eval=FALSE}
ggplot(data=mpg) + stat_smooth(aes(x=displ,
                                   y=hwy),
                               geom="line")
```

`geom="smooth"`: computed variable (standard error) is plotted

`geom="line"`: `geom_line` doesn't know how to plot standard error

# Changing the default stat

What if we want to pre-compute our statistical summaries instead of using a `ggplot2` statistical transformation?

Summarize the count of cars of each drive type:

```{r}
drive <- data.frame(type = c("four wheel",
                             "front wheel",
                             "rear wheel"),
                    count = c(sum(mpg$drv == "4"),
                              sum(mpg$drv == "f"),
                              sum(mpg$drv == "r")))
```

What does `sum()` do?

How would the code above change if there were missing values?

---

If we don't want to use `stat="count"` (the default), then we need to change it so no statistical transformation takes place.

```{r drv}
ggplot(drive) + geom_bar(mapping=aes(x=type, y=count),
                              stat="identity")
```

---

How could we change the code below to plot proportions instead?

```{r, eval=FALSE}
drive <- data.frame(type = c("four wheel",
                             "front wheel",
                             "rear wheel"),
                    prop = c(sum(mpg$drv == "4") / nrow(mpg),
                              mean(mpg$drv == "f"),
                              mean(mpg$drv == "r")))

ggplot(drive) + geom_col(mapping=aes(x=type, y=prop))
```

Hint: `nrow()` gives the number of rows in a `data.frame`

*Bonus: Can you simplify the code by finding a 'geom' function that uses geom="bar" with stat="identity" by default?*

# Derived variables

What if we only want to distinguish cars with less than 6 cylinders and cars with 6 cylinders or more?

```{r calc-4}
ggplot(data=mpg) + geom_point(aes(x=displ,
                                  y=hwy,
                                  color=cyl >= 6))
```

----

What if we want to plot highway vs city mileage while distinguishing whether each car has an engine size smaller or larger than 3 L?

```{r calc-5}
ggplot(data=mpg) + geom_point(aes(x=cty,
                                  y=hwy,
                                  color=displ >= 3))
```

----

What if we want to make a boxplot of highway mileage for cars with engine sizes smaller and larger than 3 L?


```{r calc-6}
ggplot(data=mpg) + geom_boxplot(aes(x=displ >= 3,
                                    y=hwy))
```

## Convert continous to categorical

Suppose we want to plot engine size vs highway mileage while visualizing the number of cylinders for each car.

```{r calc-1}
ggplot(data=mpg) + geom_point(aes(x=displ,
                                  y=hwy,
                                  size=cyl))
```


This is plot is kind of busy and difficult to read though.

----

What if we mapped number of cylinders to color instead?

```{r calc-2}
ggplot(data=mpg) + geom_point(aes(x=displ,
                                  y=hwy,
                                  color=cyl))
```

This isn't what we wanted: `ggplot2` is treating `cyl` as continuous.

### Use `as.factor()`

Make the variable categorical with `as.factor()`:

```{r calc-3}
ggplot(data=mpg) + geom_point(aes(x=displ,
                                  y=hwy,
                                  color=as.character(cyl)))
```

What does `as.factor()` do? Would `as.character()` also work?

## Summary: mapping aesthetics to calculated variables

- Aesthetics can be mapped to expressions that calculate new variables

- Aesthetics can be mapped to variables calculated by stats functions

# Position adjustments

Earlier, we demonstrated a useful position adjustment: "jitter"

What are other useful position adjustments?

## Position = "stack"

```{r stack}
ggplot(mpg) + geom_bar(aes(x=drv,
                           fill=hwy >= 25),
                       position="stack")

ggplot(mpg) + geom_bar(aes(x=drv,
                           fill=hwy >= 25))
```

## Position = "dodge"

```{r dodge}
ggplot(mpg) + geom_bar(aes(x=drv,
                           fill=hwy >= 25),
                       position="dodge")
```

# Scales

Each aesthetic has a scale associated with it.

Usually, a sensible default is used and we don't have to specify it manually:

```{r}
ggplot(data=mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  scale_x_continuous() + 
  scale_y_continuous()
```

versus

```{r}
ggplot(data=mpg, aes(x=displ, y=hwy)) +
  geom_point()
```

Sometimes it is useful to set a scale manually.

## Controlling scales and legends

Suppose we want to plot both highway and city mileage

```{r scl-1}
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy), color="blue") +
  geom_point(mapping=aes(x=displ, y=cty), color="red")
```

Why is there no legend?

----

To generate a legend, an aesthetic must be mapped

```{r scl-2}
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy, color="Highway")) +
  geom_point(mapping=aes(x=displ, y=cty, color="City")) +
  labs(y="Mileage")
```

----

What if we want to use both color and shape?

```{r scl-3}
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy,
                         color="Highway",
                         shape="Highway")) +
  geom_point(mapping=aes(x=displ, y=cty,
                         color="City",
                         shape="City"))
```

But this results in two legends.

## Specifying scales manually

Sometimes it's necessary to manually specify scales:

```{r scl-4, eval=FALSE}
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy,
                         color="Highway",
                         fill="Highway",
                         shape="Highway")) +
  geom_point(mapping=aes(x=displ, y=cty,
                         color="City",
                         fill="City",
                         shape="City")) +
  scale_colour_manual(name="Mileage",
                      values=c("darkred", "darkblue")) +
  scale_shape_manual(name="Mileage",
                     values=c(21, 22))
```

- Allows you to control the legend

- Allows you to specify custom levels for each aesthetic

# Coordinate systems

Each plot uses a coordinate system.

Usually, a sensible default is used and we don't have to specify it manually:

```{r}
ggplot(data=mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  coord_cartesian()
```

versus

```{r}
ggplot(data=mpg, aes(x=displ, y=hwy)) +
  geom_point()
```

Sometimes we want to specify a different coordinate system.

## Flipped

```{r flip}
ggplot(data = mpg, aes(x = manufacturer, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```

## Polar

```{r polar}
ggplot(data = mpg, aes(x = 1, fill=class)) + 
  geom_bar() +
  coord_polar(theta="y")
```

## Maps

```{r map}
library(maps)
four_corners <- map_data("state",
                  region=c("arizona",
                           "new mexico",
                           "utah",
                           "colorado"))
ggplot(four_corners) + 
  geom_polygon(mapping=aes(x=long,
                           y=lat,
                           group=group),
               fill=NA,
               color="black") +
  coord_map()
```


# Faceting

Faceting, also called "trellis" or "lattice" graphics, use subplots to visualize different subsets of a dataset, by conditioning on the levels of a another variable in the dataset.

## Faceting by 1 categorical variable

Condition on drive type:

```{r fct-1}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  facet_wrap(~drv)
```

## Faceting by 2 categorical variables

Condition on drive type and number of cylinders:

```{r fct-2}
ggplot(mpg, aes(x=displ, y=hwy)) +
  geom_point() +
  facet_grid(cyl ~ drv)
```

## Faceting by a continuous variable

Condition on intervals of city mileage:

```{r fct-3}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cut_number(cty, 3))
```

# Themes

In addition to the traditional components of a layered grammar of graphics, `ggplot` also includes themes that allow control over the overall visual appearance of the plot.

## Notable themes

Several themes are available out-of-the-box.

Default theme:

```{r theme-default}
g <- ggplot(mpg, aes(x=displ, y=hwy, color=drv)) +
  geom_point() +
  geom_smooth()
g
```

Minimal:

```{r theme-minimal}
g + theme_minimal()
```

Dark:

```{r theme-dark}
g + theme_dark()
```

Light:

```{r theme-light}
g + theme_light()
```

Void:

```{r theme-void}
g + theme_void()
```

You can build your own theme with `theme()`.

## Adjusting annotations

Themes can useful for adjusting the non-data components of a plot.

For example, adjusting axis labels:

```{r}
ggplot(mpg, aes(x=manufacturer, y=hwy)) +
  geom_boxplot() +
  theme_minimal()
```

versus

```{r}
ggplot(mpg, aes(x=manufacturer, y=hwy)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

There are too many such options to go through, so check the documentation if you want to learn how to use themes better.

# Exporting plots with `pdf()`

Use external graphics devices like `pdf()`, `png()`, and `tiff()` to export any graphics generated by R.

```{r eval=FALSE}
pdf("mpg.pdf", width=6, height=3)
ggplot(mpg, aes(x=displ, y=hwy)) + 
  geom_point(aes(color=class)) + geom_smooth() +
  labs(title="Cars with larger engines get worse mileage",
       x="Engine size (liters)", y="Highway mileage (per gallon)")
dev.off()
```

# Summary: `ggplot2` plotting (full specification)

```{r eval=FALSE}
ggplot(data = <DATASET>,
       mapping = aes(<MAPPINGS)) +
  layer(geom = <GEOM>,
        stat = <STAT>,
        position = <POSITION>) +
  <SCALE_FUNCTION>() +
  <COORDINATE_FUNCTION>() +
  <FACET_FUNCTION>()
```

# Summary: `ggplot2` plotting (concise specification)

```{r eval=FALSE}
ggplot(data = <DATASET>,
       mapping = aes(<MAPPING>)) +
  <GEOM_FUNCTION>(stat = <STAT>,
                  position = <POSITION>) +
  <COORDINATE_FUNCTION>() +
  <FACET_FUNCTION>()
```

or

```{r eval=FALSE}
ggplot(data = <DATASET>,
       mapping = aes(<MAPPINGS)) +
  <GEOM_FUNCTION>()
```




