---
title: "Data transformation"
author: "Kylie Ariel Bemis"
date: "9/21/2020"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data transformation

In order to visualize and (later) model data, we typically need to perform data transformation.

- Subset data by rows (observations) or columns (variables)

- Reorder the rows or columns of the data

- Calculate new variables from existing variables

- Calculate summary statistics of variables

One or more of these steps are often necessary to appropriately visualize or model a dataset.

----

Below, we will discuss data transformations on __tidy__ data. That is, data that is already in a form where:

- Each variable has its own column

- Each observation has its own row

- Each value has its own cell


# `dplyr`: A grammar of data manipulation

Provides a powerful, flexible "grammar" of data manopulation.

- Identify the most important data manipulation verbs and make them easy to use from R

- Provide fast performance for in-memory data with under-the-hood C++ implementation

- Use the same interface whether the data is stored in-memory or in a database on disk

Part of the "tidyverse" with `ggplot2` and others.

## Why use `dplyr`?

The `dplyr` package (mostly) doesn't implement any functionality that is missing or impossible to perform in base R.

However, code written with `dplyr`:

- Can be more expressive, concise, and human-readable

- Can be more explicit in your intentions for data manipulation

- Can be faster (sometimes) due to C++ implementation

- Can also be used with databases on disk

- Integrates with other `tidyverse` functions and packages

I will also show you how to perform the same functionality using base R.

## Review of subsetting `data.frame`s in R

- Simplifying (returns a vector)

    - Access individual columns using `df$name`

    - Access individual columns using `df[["name"]]`

- Preserving (returns a `data.frame`)

    - Subset rows using `df[i,]`

    - Subset rows using `df[c("name1", "name2")",]`

    - Subset columns using `df[,j]`

    - Subset columns using `df[,c("name1", "name2")]`

- You can subset both rows and columns at the same time

- Subsetting by a single column is always simplifying for `data.frame`s

    - Change this behavior with `drop = FALSE`

---

```{r}
df <- data.frame(x=c(1L, 2L, 5L, 9L),
                 y=c('a', 'b', 'c', 'd'),
                 `z !`=c(1.11, 2.22, 3.33, 4.0),
                 row.names=c("Jo", "Ha", "Q", "Final"),
                 check.names=FALSE,
                 stringsAsFactors=FALSE)
df
```

---

```{r}
df$x
df[["y"]]
```

---

```{r}
df$`z !`
df[["z !"]]
```

---

```{r}
df[1:3,]
df[c("Jo", "Ha", "Q"),]
```

---

```{r}
df[,2:3]
df[,c("y","z !")]
```

---

```{r}
df[,"z !"]
df[,"z !",drop=FALSE]
```

---
  
```{r}
df[1:3, c("y", "z !")]
```

## Verbs in `dplyr`

Provides most commonly used data manipulation actions.

- `filter()` subsets data by rows/observations

- `arrange()` reorders data by rows/observations

- `select()` subsets data by columns/variables

- `mutate()` creates new columns/variables

- `summarise()` calculates summary statistics

Each can be applied over levels of a categorical variable with `group_by()`.

Each takes a `data.frame` as the first argument and outputs a new `data.frame`.

# Loading `dplyr`

```{r}
library(dplyr)
```

## Name masking and namespaces in R

Several functions are "masked" when `dplyr` is loaded.

They can still be accessed by fully qualifying their names:

- `stats::filter()`

- `stats::lag()`

----

Each package in R creates its own namespace.

R finds functions based on the order packages are loaded.

You can see this search path with `search()`:

```{r}
search()
```

----

When name conflicts occur, a warning about masked names is given.

You can always use `package::function()` to find the right one.

## Example dataset

Today we will explore the `flights` dataset also used in the homework and the **R4DS** book.

```{r}
library(nycflights13)
flights
```

Note that `flights` is actually a `tibble`, which is simply a special type of `data.frame`.

I will use `data.frame` to refer to both interchangeably.

# Subsetting rows with `filter()`

Get only flights from October.

In base R:

```{r eval=FALSE}
flights[flights$month == 10,]
```

In `dplyr`:

```{r}
filter(flights, month == 10)
```

## Why `flights$month` vs `month`?

Like `ggplot2`, `dplyr` uses non-standard evaluation to facilitate ease of interactive programming.

In base R:

```{r, eval=FALSE}
flights[flights$month == 10,]
```

In `dplyr`:

```{r, eval=FALSE}
filter(flights, month == 10)
```

This approach is useful for interactive data analysis, but can cause problems when used within user-defined functions.

We may discuss more on non-standard evaluation later in the course if there is interest.

## Anatomy of `filter()`

`filter(data, condition1, condition2, condition3, ...)`

- The first argument is the data

- The following arguments are vectorized logical expressions

- Additional arguments are joined by `&` (AND)

- Rows that evaluate to `TRUE` are kept

- Rows that evluate to `FALSE` or `NA` are dropped

## Review of logical operators in R

- Standard comparison operators: `==`, `!=`, `>`, `<`, `>=`, `<=`.

    - Remember to use `==` instead of `=` when doing comparisons
    
- `dplyr::near()` for checking floating point equality

- `dplyr::between()` is a synonym for `a <= x & x <= b`

- `&`, `|`, and `!` are vectorized AND, OR, and NOT

    - Non-vectorized versions (for `if` statements) are `&&` and `||`
    
- `%in%` checks if an element exists in a set

    - E.g., `x %in% c(a,b)` is equivalent to `x == a | x == b`

- `is.na` to check for missing values

    - Remember that `NA == NA` evaluates to `NA`

---

Get only flights from Alaska Airlines or Hawaiian Airlines.

In base R:

```{r, eval=FALSE}
flights[flights$carrier %in% c("AS", "HA"),]
```

In `dplyr`:

```{r}
filter(flights, carrier %in% c("AS", "HA"))
```

---

Get only flights from Alaska Airlines or Hawaiian Airlines.

In base R:

```{r, eval=FALSE}
flights[flights$carrier == "AS" | flights$carrier == "HA",]
```

In `dplyr`:

```{r}
filter(flights, carrier == "AS" | carrier == "HA")
```

---

Get only flights between Honolulu and JFK.

In base R:

```{r, eval=FALSE}
flights[flights$origin == "JFK" & flights$dest == "HNL",]
```

In `dplyr`:

```{r}
filter(flights, origin == "JFK" & dest == "HNL")
```

---

Get only flights between Honolulu and JFK.

In base R:

```{r, eval=FALSE}
flights[flights$origin == "JFK" & flights$dest == "HNL",]
```

In `dplyr`:

```{r}
filter(flights, origin == "JFK", dest == "HNL")
```

# Reordering rows with `arrange()`

Sort by flights that departed with least delay (most ahead of schedule):

In base R:

```{r eval=FALSE}
flights[order(flights$dep_delay),]
```

In `dplyr`:

```{r}
arrange(flights, dep_delay)
```

## Use `desc()` to sort by a variable in descending order

Sort by flights that departed with most delay:

In base R:

```{r eval=FALSE}
flights[order(flights$dep_delay, decreasing=TRUE),]
```

In `dplyr`:

```{r}
arrange(flights, desc(dep_delay))
```

# Selecting columns with `select()`

Keep only date and delay columns.

In base R:

```{r eval=FALSE}
flights[,c("year","month","day","dep_delay","arr_delay")]
```

In `dplyr`:

```{r}
select(flights, year, month, day, dep_delay, arr_delay)
```

---

Keep first 9 columns (`year` through `arr_delay`).

In base R:

```{r eval=FALSE}
flights[,1:9]
```

In `dplyr`:

```{r}
select(flights, 1:9)
```

---

Keep first 9 columns (`year` through `arr_delay`) by name.

In base R:

```{r eval=FALSE}
flights[,which(names(flights)=="year"):
          which(names(flights)=="arr_delay")]
```

In `dplyr`:

```{r}
select(flights, year:arr_delay)
```

---

Keep all columns except tail number and flight number.

In base R:

```{r eval=FALSE}
flights[,-c(which(names(flights)=="tailnum"),
          which(names(flights)=="flight"))]
```

In `dplyr`:

```{r}
select(flights, -tailnum, -flight)
```

## Useful functions for selecting column/variable names

- `dplyr::starts_with("arr")`

    - matches column names that begin with "arr"

- `dplyr::ends_with("time")`

    - matches column names that end with "time"

- `dplyr::contains("dep")`

    - matches column names that contain "dep"

- `dplyr::num_range("x", 1:3)`

    - matches "x1", "x2", and "x3"

---

Keep only columns starting with "arr".

In base R:

```{r eval=FALSE}
flights[,substr(names(flights), 1, 3) == "arr"]
```

In `dplyr`:

```{r}
select(flights, starts_with("arr"))
```

## Rename columns with `rename()`

`rename()` is a variant of `select()` that keeps all variables/columns while renaming the specified ones.

In base R:

```{r eval=FALSE}
flights2 <- flights
names(flights2)[names(flights2)=="year" |
                  names(flights2)=="month" |
                  names(flights2)=="day"] <- c("YEAR","MONTH","DAY")
```

In `dplyr`:

```{r}
rename(flights, YEAR=year, MONTH=month, DAY=day)
```

# Transforming variables with `mutate()`

Create a new variable giving the average air speed (in mph) of each flight.

In base R:

```{r eval=FALSE}
flights2 <- flights
flights2$speed <- flights2$distance / flights2$air_time * 60
```

In `dplyr`:

```{r}
mutate(flights, speed = distance / air_time * 60)
```

---

Create two new variables giving (1) the average air speed (in mph) of each flight and (2) the amount of time gained in the air.

In base R:

```{r eval=FALSE}
flights2 <- flights
flights2$speed <- flights2$distance / flights2$air_time * 60
flights2$gain <- flights2$arr_delay - flights2$dep_delay
```

In `dplyr`:

```{r}
mutate(flights,
       speed = distance / air_time * 60,
       gain = arr_delay - dep_delay)
```

## Anatomy of `mutate()`

`mutate(data, var1 = expr1, var2 = expr2, ...)`

- The first argument is the data

- The following arguments are named vectorized expressions that output a vector of the same length

- When used with `data.frame`s, you can use variables created in the same `mutate()` call in the subsequent expressions

## Useful functions for creating variables

- Arithmetic such as `+`, `-`, `*`, `/`, `^`, etc.

    - These are vectorized and will recycle shorter variables
    
- Modular arithmetic such as `%%` and `%/%`

    - Useful for breaking apart integers (e.g., time into hours + minutes)

- Logs such as `log()`, `log2()` and `log10`

    - Useful for data with multiplicative variance
    
- Offsets such as `dplyr::lead()` and `dplyr::lag()`

    - Useful for running differences and data over time points
    
- Cumulative summaries such as `cumsum()`, `cumprod()`, `cummax()`, and `cummin()`

    - Also `dplyr::cummean()` for running means
    
- Logical operators such as ``==`, `!=`, `>`, `<`, `>=`, `<=`, etc.

    - Useful for turning continuous variables into categorical
    
- `dplyr::n()` gives the number of observations

    - `n()` can only be used inside `mutate()`, `filter()` and `summarise()`

## New variables with `transmute()`

`transmute()` is a variant of `mutate()` that keeps only the new variables and drops the rest.

In base R:

```{r eval=FALSE}
data.frame(distance = flights$distance,
           speed = flights$distance / flights$air_time * 60,
           gain = flights$arr_delay - flights$dep_delay,
           gain_per_mile = (flights$arr_delay - flights$dep_delay) /
             flights$distance)
```

In `dplyr`:

```{r}
transmute(flights,
          distance = distance,
          speed = distance / air_time * 60,
          gain = arr_delay - dep_delay,
          gain_per_mile = gain / distance)
```

# Summary statistics with `summarise()`

Get the mean departure and arrival delay.

In base R:

```{r eval=FALSE}
data.frame(mean_dep_delay = mean(flights$dep_delay,
                                 na.rm=TRUE),
           mean_arr_delay = mean(flights$arr_delay,
                                 na.rm=TRUE))
```

In `dplyr`:

```{r}
summarise(flights,
          mean_dep_delay = mean(dep_delay, na.rm=TRUE),
          mean_arr_delay = mean(arr_delay, na.rm=TRUE))
```

## Anatomy of `summarise()`

`summarise(data, summary1 = expr1, summary2 = expr2, ...)`

- The first argument is the data

- The following arguments are expressions that output a single value from a vector of values

- It is particularly important to consider missing values when summarizing data

- Also available as `summarize()`

## Useful functions for calculating summary statistics

- Measures of location such as `mean()` and `median()`

- Measures of spread such as `sd()`, `var()`, `IQR()`, and `mad()`

- Measures of rank such as `min()`, `max()`, and `quantile()`

- Counts such as:

    - `dplyr::n()` gives the number of observations
    
    - `sum(!is.na(x))` gives the number of non-missing values
    
    - `dplyr::n_distinct()` gives the number of unique values
    
- Remember that `sum(x == 10)` gives the count of `x == 10`

    - What does `mean(x == 10)` calculate?

---

Calculate the proportion of flights delayed more than 2 hours on arrival.

```{r}
summarise(flights, mean(arr_delay > 120, na.rm=TRUE))
```

----

Calculate the number of unique airline carriers.

```{r}
summarise(flights, n_distinct(carrier))
```

----

Calculate the proportion of flights with missing air times.

```{r}
summarise(flights, sum(is.na(air_time)) / n())
```

## Grouped transformations with `group_by()`

`summarise()` and the other data manipulation verbs in `dplyr` become much more powerful when paired with `group_by()`.

Count the number of flights from each carrier.

```{r}
summarise(group_by(flights, carrier), n())
```

Calculate the average arrival delay for each carrier.

```{r}
summarise(group_by(flights, carrier),
          mean(arr_delay, na.rm=TRUE))
```

You can group by multiple variables.

Use `ungroup()` to ungroup a grouped dataset.

# Piping with the pipe operator

Combining multiple `dplyr` verbs becomes much more expressive when used with the pipe operator `%>%`.

The pipe operator takes the return value of the expression on the LHS and turns it into the first argument of the function on the RHS.

```{r eval=FALSE}
foo(bar(baz(x)))
```

is the same as

```{r eval=FALSE}
baz(x) %>% bar() %>% foo()
```

is the same as

```{r eval=FALSE}
x %>% baz() %>% bar() %>% foo()
```

----

```{r eval=FALSE}
summarise(group_by(flights, carrier),
          mean(arr_delay, na.rm=TRUE))
```

becomes

```{r eval=FALSE}
group_by(flights, carrier) %>%
          summarise(mean(arr_delay, na.rm=TRUE))
```

or

```{r}
flights %>%
  group_by(carrier) %>%
  summarise(mean(arr_delay, na.rm=TRUE))
```

# Data transformation + visualization

You can chain together `dplyr` verbs with `ggplot2` too.

Visualize the average delay for each carrier.

```{r}
library(ggplot2)
flights %>%
  group_by(carrier) %>%
  summarise(delay=mean(arr_delay, na.rm=TRUE)) %>%
  ggplot(aes(x=carrier, y=delay)) + geom_col()
```

---

For each destination, visualize the proportion of delayed arriving flights versus the average flight time in the air.

```{r}
flights %>%
  group_by(dest) %>%
  summarise(prop_delayed = mean(arr_delay > 0, na.rm=TRUE),
            flight_time = mean(air_time, na.rm=TRUE),
            count = n()) %>%
  ggplot(aes(x=flight_time, y=prop_delayed)) + geom_point()
```

---

Map the number of flights from each destination to an aesthetic.

Add a smooth fitted line.

```{r}
flights %>%
  group_by(dest) %>%
  summarise(prop_delayed = mean(arr_delay > 0, na.rm=TRUE),
            flight_time = mean(air_time, na.rm=TRUE),
            count = n()) %>%
  ggplot(aes(x=flight_time,
             y=prop_delayed)) + 
  geom_point(aes(size=count), alpha=1/2) +
  geom_smooth()
```

# Practice at home

- For carriers that flew more than 1000 flights in 2013, find the number of flights that weren't delayed on arrival.

- Find the average distance flown by each carrier in each month of 2013.

- Plot to total miles flown each month in 2013.

- Plot the proportion of flights delayed by 10 minutes or more for each hour of the day.

- Plot the total distance flown versus the total time in arrival delays for each plane.

- Plot the relationship between the total time in the air and the total distance flown for each plane.

- Plot the average speed flown versus the average distance flown for each destination.

- Find the fastest plane.


