---
title: "Relational Data and SQL"
author: "Kylie Ariel Bemis"
date: "10/5/2020"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R and databases

Often, relational data is stored in a database managed by an RDBMS, so we need to know how to work with data in a database.

A database can also be a good solution for data which you do not want to load all into memory at once.

R and `dplyr` can interface with databases through the `DBI` package and a suitable backend:

- `RSQLite` works with SQLite databases

- `RMySQL` works with MySQL databases

- `RPostgreSQL` works with PostgreSQL databases

etc.

We can work with these with `dplyr` using the `dbplyr` package.

## R and databases with `dbplyr`

Suppose we wish to work with an SQLite database.

First we need to install the necessary packages.

```{r eval=FALSE}
install.packages(c("DBI", "RSQLite", "dbplyr"))
```

Then we create some example tables.

```{r}
library(tidyverse)
library(RSQLite)

Student <- tribble(
  ~sid, ~name, ~major, ~gpa,
  "0001", "John", "CS", NA,
  "0002", "Lucy", "DS", 4.00,
  "0003", "Aiden", "CS", 3.33
)

Course <- tribble(
  ~crn, ~course, ~dept, ~term,
  "00234", "Intro CS", "CS", "Fall2020",
  "00653", "Intro DS", "CS", "Fall2020",
  "00783", "Algorithms", "CS", "Fall2020",
  "01945", "ML & AI", "EE", "Spring2021"
)

Enrolled <- tribble(
  ~sid, ~crn, ~grade,
  "0002", "00653", "A",
  "0002", "01945", NA,
  "0003", "00783", "B+"
)
```

## Using a SQLite database in-memory

We can create an ad-hoc SQLite database in memory.

We would replace the second argument of `dbConnect()` with a filepath to connect to a file-based database.

```{r}
con <- dbConnect(SQLite(), ":memory:")

dbWriteTable(con, "student", Student)
dbWriteTable(con, "course", Course)
dbWriteTable(con, "enrolled", Enrolled)
```

Use `tbl()` to create a database-backed data frame.

```{r}
student <- tbl(con, "student")
course <- tbl(con, "course")
enrolled <- tbl(con, "enrolled")
```

# Relational algebra

## Projection

Projection selects attributes (columns).

SQL:

```{r}
dbGetQuery(con,
           "SELECT name, major
           FROM student")
```

dplyr:

```{r}
select(student, name, major)
```

---

SQL:

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student")
```

dplyr:

```{r}
select(student, everything())
```

## Selection

Selection filters tuples (rows) based on conditions.

SQL:

```{r}
dbGetQuery(con,
           "SELECT name, gpa
           FROM student
           WHERE gpa > 3.5")
```

dplyr:

```{r}
student %>%
  select(name, gpa) %>%
  filter(gpa > 3.5)
```

---

```{r}
dbGetQuery(con,
           "SELECT name, gpa
           FROM student
           WHERE gpa IS NOT NULL")
```


```{r}
student %>%
  select(name, gpa) %>%
  filter(!is.na(gpa))
```

## Rename

Rename attributes (columns).

SQL:

```{r}
dbGetQuery(con,
           "SELECT
            sid AS ID,
            name AS student
           FROM student")
```

dplyr:


```{r}
student %>%
  select(sid, name) %>%
  rename(ID=sid, student=name)
```

```{r}
transmute(student, ID=sid, student=name)
```

## Cross-product

Take all combinations of all tuples (rows) in both tables.

SQL:

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student, enrolled")
```

dplyr:

No obvious form; rarely used.

## Natural join

Combinations of tuples (rows) where they have equal values for common attributes (columns).

SQL:

```{r}
dbGetQuery(con,
           "SELECT DISTINCT
            S.sid, name, major, gpa, crn, grade
           FROM student S, enrolled E
           WHERE S.sid = E.sid")
```

dplyr:

```{r}
inner_join(student, enrolled)
```

## Set operations

Same as set theory equivalents. Both tables must have the same attributes.

```{r}
Student2 <- tribble(
  ~sid, ~name, ~major, ~gpa,
  "0002", "Lucy", "DS", 4.00,
  "0004", "Rosa", "DS", 3.67
)

dbWriteTable(con, "student2", Student2)

student2 <- tbl(con, "student2")

student

student2
```

SQL:

```{r}
# union
dbGetQuery(con,
           "SELECT * FROM student
           UNION
           SELECT * FROM student2")

# difference
dbGetQuery(con,
           "SELECT * FROM student
           EXCEPT
           SELECT * FROM student2")

# intersection
dbGetQuery(con,
           "SELECT * FROM student
           INTERSECT
           SELECT * FROM student2")
```

dplyr:

```{r}
union(student, student2)

setdiff(student, student2)

intersect(student, student2)
```

# Aggregation

## Simple aggregation

SQL:

```{r}
dbGetQuery(con,
           "SELECT AVG(gpa)
           FROM student")
```

dplyr:

```{r}
summarize(student, mean(gpa, na.rm=TRUE))
```

## Simple aggregation with filtering

SQL:

```{r}
dbGetQuery(con,
           "SELECT AVG(gpa)
           FROM student
           WHERE major = 'CS'")
```

dplyr:

```{r}
student %>%
  filter(major == "CS") %>%
  summarize(mean(gpa, na.rm=TRUE))
```

## Grouped aggregation

SQL:

```{r}
dbGetQuery(con,
           "SELECT major, AVG(gpa)
           FROM student
           GROUP BY major")
```

dplyr:

```{r}
student %>%
  group_by(major) %>%
  summarize(mean(gpa, na.rm=TRUE))
```

## Grouped aggregation with filtering

SQL:

```{r}
dbGetQuery(con,
           "SELECT major, AVG(gpa) AS gpa
           FROM student
           GROUP BY major
           HAVING gpa > 3.5")
```

dplyr:

```{r}
student %>%
  group_by(major) %>%
  summarize(gpa=mean(gpa, na.rm=TRUE)) %>%
  filter(gpa > 3.5)
```

## Grouped aggregation with sorting

SQL:

```{r}
dbGetQuery(con,
           "SELECT major, AVG(gpa) AS gpa
           FROM student
           GROUP BY major
           ORDER BY gpa DESC")
```

dplyr:

```{r}
student %>%
  group_by(major) %>%
  summarize(gpa=mean(gpa, na.rm=TRUE)) %>%
  arrange(desc(gpa))
```

# Joins

```{r}
Enrolled <- tribble(
  ~student, ~course, ~grade,
  "0002", "00653", "A",
  "0002", "01945", NA,
  "0003", "00783", "B+"
)

dbWriteTable(con, "enrolled", Enrolled, overwrite=TRUE)

enrolled <- tbl(con, "enrolled")
```

## A simple join

There are often multiple ways to express the same join in SQL.

```{r}
dbGetQuery(con,
           "SELECT
           sid, name, course, grade
           FROM student, enrolled
           WHERE sid = student")
```

```{r}
dbGetQuery(con,
           "SELECT
           sid, name, course, grade
           FROM student
           JOIN enrolled
           ON sid = student")
```

## Inner join

Keep only rows with matching key-values in both tables.

SQL:

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student
           JOIN enrolled
           ON sid = student")
```

dplyr:

```{r}
inner_join(student, enrolled, by=c("sid"="student"))
```

## Left join

Keep all rows in left-hand table.

SQL:

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student
           LEFT JOIN enrolled
           ON sid = student")
```

dplyr:

```{r}
left_join(student, enrolled, by=c("sid"="student"))
```

## Left join with filtering and sorting

SQL:

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student
           LEFT JOIN enrolled
           ON sid = student
           WHERE major = 'CS'
           ORDER BY name")
```

dplyr:

```{r}
student %>%
  left_join(enrolled, by=c("sid"="student")) %>%
  filter(major == "CS") %>%
  arrange(name)
```

## Semi join

Keep rows based on matches in another table.

SQL:

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student
           WHERE sid IN
           (
            SELECT student
            FROM enrolled
           )")
```

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student
           WHERE EXISTS
           (
            SELECT 1
            FROM enrolled
            WHERE student.sid = enrolled.student
           )")
```

dplyr:

```{r}
semi_join(student, enrolled, by=c("sid"="student"))
```

## Anti join

Drop rows based on matches in another table.

SQL:

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student
           WHERE sid NOT IN
           (
            SELECT student
            FROM enrolled
           )")
```

```{r}
dbGetQuery(con,
           "SELECT *
           FROM student
           WHERE NOT EXISTS
           (
            SELECT 1
            FROM enrolled
            WHERE student.sid = enrolled.student
           )")
```

dplyr:

```{r}
anti_join(student, enrolled, by=c("sid"="student"))
```

# SQL and `dplyr`

## Comparison

Below is a table of common data manipulation tasks in both SQL and `dplyr`.

| SQL | dplyr | task |
|-----|-------|------|
| SELECT | `select()` | select columns |
| DISTINCT | `distinct()` | eliminate duplicates |
| AS | `mutate()` | rename columns |
| FROM | - | specify tables |
| WHERE | `filter()` | filter rows |
| GROUP BY | `group_by()` | group values |
| HAVING | `filter()` | filter groups |
| ORDER BY | `arrange()` | sort table |


## Lazy queries

When `dplyr` is used with databases, it translates R code into SQL queries.

We can see the query it will execute with `show_query()`

```{r}
student %>%
  left_join(enrolled, by=c("sid"="student")) %>%
  filter(major == "CS") %>%
  arrange(name) %>%
  show_query()
```

(Sometimes the translation isn't quite the same query we'd write ourselves.)

The query is only executed by the database when the results are needed.

This means that the intermediate results aren't stored a standard data frame, but a "lazy data frame", so some functions may not know how to use it.

## Collecting results

To force results to be collected into a standard data frame, use `collect()`.

```{r}
student %>%
  left_join(enrolled, by=c("sid"="student")) %>%
  filter(major == "CS") %>%
  arrange(name) %>%
  collect()
```


