---
title: "DS5110 Homework 6"
author: "Ameya Santosh Gidh"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part A

## Problem 1

```{r}
# Suppress startup messages from packages
suppressPackageStartupMessages(library(tidyverse))

# Load necessary libraries
library(readr)      # For reading data
library(dplyr)      # For data manipulation
library(ggplot2)    # For data visualization
library(tidyverse)  # Comprehensive data manipulation and visualization package
library(glmnet)     # For fitting Lasso and Elastic-Net regularized generalized linear models
library(stringr)    # For string manipulation
library(tokenizers) # For tokenization
library(lubridate)  # For working with dates
library(tidytext)   # For text mining with tidy data principles

# Check if tidyverse package is installed, if not, install it
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

```


```{r}
dir <- "./"
path <- file.path(dir, "twitter", "realDonaldTrump-20201106.csv")
trump <- read_csv(path, col_types=cols(id=col_character()))
more_stop_words <- c("realdonaldtrump", "donaldtrump",
"donald", "trump", "rt")
trump_tidy <- trump %>%
filter(!isRetweet, str_detect(text, "[:space:]")) %>%
select(id, text, retweets, date) %>%
unnest_tokens("word", text, token="regex") %>%
mutate(word=str_replace(word, "ˆ#", "HASHTAG"),
word=str_replace(word, "ˆ@", "AT"),
word=str_replace_all(word, "[[:punct:]]+", ""),
word=str_replace(word, "ˆHASHTAG", "#"),
word=str_replace(word, "ˆAT", "@")) %>%
anti_join(stop_words, by="word") %>%
filter(!word %in% more_stop_words,
!str_detect(word, "@"),
!str_detect(word, "amp"),
!str_detect(word, "http"),
str_length(word) > 0L)
```

visualize the top 20 most common terms.
```{r}
trump_tidy %>%
count(word) %>%
top_n(20) %>%
ggplot(aes(x=reorder(word, n), y=n)) +
geom_col() +
coord_flip() +
labs(x="Term", y="Count",
title="Top terms in Donald Trump's tweets") +
theme_minimal()
```
The visualization above indicates that among the top 20 tweets by Donald Trump, the most frequent theme is related to the term "President," with other tweets also revolving around political topics.

## Problem 2

#### Donald Trump’s tweets for each year from 2015-2020
```{r}
library(lubridate)
trump_tidy %>%
mutate(year=year(date)) %>%
filter(year %in% 2015:2020) %>%
count(word, year) %>%
group_by(year) %>%
top_n(10) %>%
ggplot(aes(x=reorder_within(word, n, year),
y=n, fill=factor(year))) +
geom_col(show.legend=FALSE) +
coord_flip() +
scale_x_reordered() +
scale_y_continuous(labels=NULL) +
scale_fill_brewer(palette="Dark2") +
facet_wrap(~year, scales="free", ncol=2) +
labs(x="Term", y="Count",
title="Top terms in Trump's tweets by year") +
theme_minimal()
```


The plot reveals that before the election, Donald Trump's tweets mainly focused on topics such as Make America Great Again and Hillary Clinton. After winning the election, his tweets shifted towards discussing China, jobs, and Democrats. Interestingly, prior to the election, we observe the resurgence of Make America Great Again along with mentions of Joe Biden, who eventually became President.
We can see Trump’s campaign hashtags in 2015 and 2016 such as #Trump2016 and #MakeAmericaGreatAgain.
“Fake” is a recurring term from 2017-2020. Biden appears in 2020 due to the election.

# Part B

## Problem 3

```{r}
trump_year <- trump_tidy %>%
mutate(year=year(date)) %>%
filter(year %in% 2015:2020) %>%
count(word, year) %>%
bind_tf_idf(word, year, n)
trump_year %>%
group_by(year) %>%
top_n(10, tf_idf) %>%
ggplot(aes(x=reorder_within(word, tf_idf, year),
y=tf_idf, fill=factor(year))) +
geom_col(show.legend=FALSE) +
coord_flip() +
scale_x_reordered() +
scale_y_continuous(labels=NULL) +
scale_fill_brewer(palette="Dark2") +
facet_wrap(~year, scales="free", ncol=2) +
labs(x="Term", y="tf-idf",
title="Each year's characteristic terms from Trump's tweets") +
  theme_minimal()
```


It came as no surprise that "Make America Great Again" and "Trump2016" were the most commonly tweeted expressions, considering that 2015-2016 was characterized by election rallies. In the following years, Trump often tweeted about his policies, alongside regularly denouncing "Fake News." As expected, COVID and Coronavirus were the primary topics of Trump's discussions in 2020, which was to be anticipated.
Using tf-idf, we now see coronavirus and COVID appear to distinguish the 2020 tweets. We can also see that
2015-2017 are characterized by hashtags while 2018-2020 use fewer unique hashtags

## Problem 4

```{r}
trump_dtm <- trump_tidy %>%
filter(year(date) >= 2016) %>%
count(id, word) %>%
cast_sparse(id, word, n)
dim(trump_dtm)

```

```{r}
trump_dtm[1:6, 1:6]
```

We extract the tweet IDs to get the target variable (# of retweets) in the correct order.
```{r}
trump_ids <- tibble(id=rownames(trump_dtm))
trump_ids <- left_join(trump_ids, trump)
trump_ids

```

```{r}
trump_rt <- trump_ids$retweets
```



```{r}
set.seed(2020)
fit_cv <- cv.glmnet(trump_dt, trump_rt)

```

```{r}
plot(fit_cv)
```

```{r}
fit_cv
```
The best model based on MSE selects lambda = 181.7, which uses 689 terms in the model.
The most sparse model within 1 standard error of the best model’s MSE has lambda = 608.9, and uses only
59 terms in the model

## Problem 5

```{r}
trump_beta <- coef(fit_cv, s="lambda.min")
trump_beta <- tibble(word=rownames(trump_beta),
coef=as.numeric(trump_beta))
trump_beta %>%
top_n(15) %>%
ggplot(aes(x=reorder(word, coef), y=coef)) +
geom_col() +
coord_flip() +
labs(x="Term", y="Regression Coefficient") +
theme_minimal()

```
We can see the top term for determining re-tweets is #FNN, presumably referring to Fox News Network.
The Trump-ism “covfefe” makes an appearance for correlating with a high number of re-tweets.
We can find the context for some of these terms by searching for them in the original tweets.


```{r}
trump_lower <- trump %>%
filter(year(date) >= 2016) %>%
mutate(text=str_to_lower(text))
```

```{r}
trump_lower %>%
filter(str_detect(text, "#fnn")) %>%
select(date, text, retweets)
```


```{r}
trump_lower %>%
filter(str_detect(text, "insult")) %>%
select(date, text, retweets)

```


```{r}
trump_lower %>%
filter(str_detect(text, "starved")) %>%
select(date, text, retweets)
```


```{r}
trump_lower %>%
filter(str_detect(text, "weli")) %>%
select(date, text, retweets)
```


```{r}
trump_lower %>%
filter(str_detect(text, "quarantine")) %>%
select(date, text, retweets)
```

```{r}
trump_lower %>%
filter(str_detect(text, "covfefe")) %>%
select(date, text, retweets)
```

```{r}
trump_lower %>%
filter(str_detect(text, "way[:punct:]+and")) %>%
select(date, text, retweets)
```



```{r}
trump_lower %>%
filter(str_detect(text, "amazing[:punct:]+tremendous")) %>%
select(date, text, retweets)
```

